\documentclass{article}

\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[capitalize,noabbrev]{cleveref}

% Blind workshop submission. If accepted, switch to:
% \usepackage[accepted]{icml2026}
\usepackage{icml2026}

\newcommand{\zscore}{z = (x-\mu)/\sigma}

\icmltitlerunning{The Geometry of Relativity}

\begin{document}

\twocolumn[
  \icmltitle{The Geometry of Relativity: Context-Normalized Encoding of Gradable Adjectives in LLMs}

  \begin{icmlauthorlist}
    \icmlauthor{Anonymous Authors}{anon}
  \end{icmlauthorlist}

  \icmlaffiliation{anon}{Anonymous Institution}
  \icmlcorrespondingauthor{Anonymous Authors}{anonymous@example.com}
  \icmlkeywords{mechanistic interpretability, representation geometry, language models, gradable adjectives}

  \vskip 0.3in
]

\printAffiliationsAndNotice{}

\begin{abstract}
Large language models make context-dependent scalar judgments: the same value can be ``tall'', ``short'', or ordinary depending on the comparison class in the prompt. We study whether Gemma 2 represents these judgments using raw magnitudes or context-normalized standing. Across gradable adjective pairs, dense behavioral grids and residual-stream probes show a strong signal aligned with \zscore. Causal steering and cross-pair transfer suggest a partly shared direction, but lexical and output-facing components explain a substantial fraction of the effect. We argue for a mixed mechanism: a transferable residual component carries context-normalized standing, while pair-specific lexical circuitry shapes the final judgment.
\end{abstract}

\section{Introduction}

TODO: One paragraph on relative scalar meaning. Example: 180cm is tall in one context and ordinary in another.

TODO: One paragraph on why this is a mechanistic interpretability question, not only a behavioral pragmatics question.

TODO: State the claim conservatively: Gemma 2 contains a partly shared context-normalized direction, but the mechanism is mixed.

\paragraph{Contributions.}
TODO: Convert the current seven evidence lines from \texttt{docs/paper\_outline.md} into 4--6 reviewer-readable bullets.

\section{Related Work}

TODO: Cover gradable adjectives and pragmatic thresholding, numerical representations, linear representation/probing, activation steering, and SAE feature decomposition.

TODO: Key distinction: prior work shows behavioral/contextual sensitivity or numeric representations; this paper tests where and how a context-normalized scalar variable is represented and used.

\section{Experimental Setup}

\paragraph{Models.}
TODO: Gemma 2 2B and 9B, activation sites, layers, and tokenizer details.

\paragraph{Tasks.}
TODO: Relative adjective pairs plus absolute control; prompt construction; dense x-by-z grid; seeds.

\paragraph{Measurements.}
TODO: Logit-difference target, residual activations, probe construction, steering protocol, SAE audit.

\section{Behavioral Evidence for Context-Normalized Judgments}

TODO: Main figure/table for dense-grid behavioral correlation with \zscore.

TODO: Make explicit which result rules out raw magnitude as the sole explanation.

\section{Representation Geometry}

TODO: PCA and probe results. Use the careful framing: PC1 is z-like for many pairs, with known exceptions.

TODO: Include shared direction alignment and cross-pair transfer evidence.

\section{Causal Use and Lexical Decomposition}

TODO: Encode-early/use-late result from V12.

TODO: Lexical projection versus residual direction from V12.1/V12.2.

TODO: State the mixed-mechanism interpretation plainly.

\section{Sparse Feature Evidence}

TODO: SAE z-correlated features, raw-number/token controls, and lexical audit.

TODO: Avoid claiming pure relative-standing features.

\section{Limitations}

TODO: Include lexical confounding, prompt-template sensitivity, single model family, limited causal localization, and correlation/steering not being a full circuit proof.

\section{Conclusion}

TODO: One paragraph. The core contribution is a mechanistic case study of context-normalized scalar meaning, not a complete reverse-engineering of gradable adjective semantics.

\bibliography{references}
\bibliographystyle{icml2026}

\appendix

\section{Appendix}

TODO: Full prompt templates, hyperparameters, all ablations, additional figures, and source paths for headline numbers.

\end{document}
